### 2019-01-27 metrics2.tex
Metrcis

\begin{center}
	Econometrics 1. Problem Set 2.
\end{center}
\section*{Problem 1}
Suppose that the population model determining $y$ is
\begin{align*}
y = \beta_0 + \beta_1 x_1 + \beta_2x_2 + \beta_3x_3 + u
\end{align*}
and this model satisfies the Gauss-Markov assumptions. However, we estimate the model that
omits $x_3$. Let $\tilde{\beta_1}, \tilde{\beta_2}$ be the OLS estimators from the regression of $y$ on $x_1$ and $x_2$. Show
that the expected value of $\tilde{\beta_1}$ (given the values of the independent variables in the sample) is
\begin{align*}
\expect \tilde{\beta_1} = \beta_1 + \beta_3 \frac{\sum_{i=1}^n \hat{r_{i1}x_{i3}}}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align*}
where the $\hat{r_{i1}}$ are the OLS residuals from the regression of $x_1$ on $x_2$.

\textbf{Solution}
Plugging $y_i = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2}+\beta_3x_{i3}$ into
\begin{align*}
\tilde{\beta_1} = \frac{\sum_{i=1}^n \hat{r_{i1}}y_i}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align*}
one can obtain:
\begin{align} \label{eq1}
\tilde{\beta_1} = \frac{\sum_{i=1}^n \hat{r_{i1}}(\beta_0 + \beta_1x_{i1} + \beta_2x_{i2}+\beta_3x_{i3} + u_i)}{\sum_{i=1}^n \hat{r_{i1}}^2} = \nonumber \\
=\frac{\beta_0 \sum_{i=1}^n \hat{r_{i1}} + \beta_1 \sum_{i=1}^n \hat{r_{i1}}x_{i1} + \beta_2 \sum_{i=1}^n \hat{r_{i1}}x_{i2} + \beta_2 \sum_{i=1}^n \hat{r_{i1}}x_{i2} + \sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align}
By properties of OLS estimators (corollary from FOC): $\sum_{i=1}^n \hat{r_{i1}} = 0, \sum_{i=1}^n \hat{r_{i1}}x_{i2} = 0$. Moreover,
\begin{align*}
\sum_{i=1}^n \hat{r_{i1}}x_{i1} = \sum_{i=1}^n \hat{r_{i1}} (\hat{\alpha_0} + \hat{\alpha_1}x_{i2} + \hat{r_{i1}}) = 0 + 0 + \sum_{i=1}^n \hat{r_{i1}}^2
\end{align*}
where $\hat{\alpha_0}, \hat{\alpha_1}$ are OLS estimates for coefficients of regression of $x_1$ on $x_2$. Plugging it into (\ref{eq1}):
\begin{align*}
\tilde{\beta_1} = \beta_1 + \beta_3\frac{\sum_{i=1}^n\hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2} + \frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2}\\
\expect \tilde{\beta_1} = \beta_1 + \beta_3\frac{\sum_{i=1}^n\hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2} + \expect\left[ \frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2} \right]
\end{align*}
\begin{align*}
\hat{r_{i1}} = x_{i1} - \hat{x_{i1}} = x_{i1} - \hat{\alpha_0} - \hat{\alpha_1}x_{i2}\ \to \expect[f(\hat{r_{i1}})|x_{i1}, x_{i2}] = f(\hat{r_{i1}}) \ \forall\ f(\cdot)
\end{align*}
Then,
\begin{align*}
\expect\left[\frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2} \right] = \expect\left[\expect\left[\frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2}\bigg|x_{1}, x_{2}\right] \right] \underset{\text{i.i.d.}}{\propto} \expect\left[\sum_{i=1}^n\hat{r_{i1}}\expect [u_i|x_{i1}, x_{i2}]\right] \underset{\text{G.-M. assump.}}{=} 0 
\end{align*}
Thus, 
\begin{align*}
\expect \tilde{\beta_1} = \beta_1 + \beta_3\frac{\sum_{i=1}^n\hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align*}
\section*{Problem 2}
\begin{enumerate}[(i)]
	\item Consider the simple regression model $y = \beta_0 +\beta_1x+u$ under the first four Gauss-Markov
	assumptions. For some function $g(x)$, for example $g(x) = x^2$ or $g(x) = \ln (1 + x^2)$, define
	$z_i = g(x_i)$. Define a slope estimator as
	\begin{align*}
	\tilde{\beta_1} = \frac{\sum_{i=1}^n (z_i - \bar{z})y_i}{\sum_{i=1}^n (z_i - \bar{z})x_i}
	\end{align*}
	Show that $\tilde{\beta_1}$ is linear and unbiased.
	\item Add the homoskedasticity assumption, MLR.5. Show that
	\begin{align*}
	Var \tilde{\beta_1} = \sigma^2 \frac{\sum_{i=1}^n (z_i -  \bar{z})^2}{\left(\sum_{i=1}^n (z_i - \bar{z})x_i\right)^2}
	\end{align*}
	\item Show directly that, under the Gauss-Markov assumptions, $Var\hat{\beta_1} \le Var\tilde{\beta_1}$, where $\hat{\beta_1}$ is the OLS estimator.
\end{enumerate}

@@

\begin{enumerate}[(a)]
	\item First of all $x_i \neq 0\ \forall i$ then the estimator is well-defined.
	\item \begin{align*}
	\expect[\tilde{\gamma}|x_1, \dots, x_n] = \expect\left[\frac{\summa (\gamma x_i + \varepsilon_i)x_i^3}{\summa x_i^4}\bigg|x_1, \dots, x_n\right] &= \gamma + \frac{1}{\summa x_i^4} \summa x_i^3\expect[\varepsilon_i|x_1, \dots, x_n] =\\
	 &\underset{\text{G.-M. assump}}{=} \gamma + 0
	\end{align*}
	\item \begin{align*}
	Var(\tilde{\gamma}|x_1, \dots, x_n) = Var\left[\frac{\summa (\gamma x_i + \varepsilon_i)x_i^3}{\summa x_i^4}\bigg|x_1, \dots, x_n\right] \underset{\text{i.i.d}}{=} \frac{\summa x_i^6 Var[\varepsilon_i|x_i]}{\left(\summa x_i^4\right)^2} 
	\end{align*}
	If $Var(\varepsilon|x) = const = \sigma^2$ then
	\begin{align*}
	Var(\tilde{\gamma}|x_1, \dots, x_2) = \sigma^2 \frac{\summa x_i^6}{\left(\summa x_i^4\right)^2} 
	\end{align*}
\end{enumerate}
\section*{Problem 5}
In this problem set you continue to explore the returns to education and the gender gap
in earnings. The data set is the same as for problem set 1 (cps99 ps1.csv; see problem set 1
for a description of the variables). Consider the regression of average hourly earnings (\textit{ahe}) on
years of education (\textit{yrseduc}). Consider the following three variables that have been omitted
from the regression:
\begin{enumerate}[(a)]
\item gender
\item a binary variable that = 1 if the worker’s last name falls in the first half of the alphabet,
and = 0 if it falls in the second half.
\item the worker’s native ability (for example IQ or some better measure)
\end{enumerate}
For each: Will this omission arguably lead to omitted variable bias? Why or why not?
If your answer is that it will, state the sign of

@@

\textbf{Solution}
\begin{enumerate}[(a)]
	\item An idealized randomized controlled experiment should be as follows: we should absolutely randomly choose two groups of people. Absolutely randomly in the sense that each person of the USA could equally likely be chosen for this experiment. People of the first group should be given an additional year of education, while people from the second group should be abstained from the education over the year. Then after this year 
\end{enumerate}
\section*{Problem 7}
Estimate two regressions: (i) \textit{ahe} on \textit{yrseduc} and (ii) \textit{ahe} on \textit{yrseduc} and \textit{female}.
\begin{enumerate}[(a)]
\item In regression (ii), what is the coefficient on \yrseduc? Explain what this means.
\item In regression (ii), test the hypothesis that the population coefficient on female in
specification (ii) is zero, against the hypothesis that it is nonzero, at the 5\% significance level.
In everyday words (not statistical terms), what precisely is the hypothesis is that you are
testing?
\item Does the coefficient on \textit{yrseduc} change from regressions (i) to (ii) in a substantively
important way, that is, is the difference between the two estimates large in a real-world sense?
What is the reason that including \textit{female} in the regression results in a large (or only a small)
change in the coefficient on \yrseduc? (Hint: what is the correlation between \yrseduc\ and
\textit{female}?)

@@

\begin{enumerate}[(a)]
	\item Coefficient on \yrseduc\ is equal to 1.34147. That means that holding a gender of the person fixed each additional year of education on average gives 1.34147 dollars of increase of the per hour earnings.
	\item Using coeftest() it can be obtained that the null hypothesis should be rejected on the 5\% significance level. The hypothesis was that holding a number of years of education fixed both males and females get on average equal salary per hour.
	\item Coefficient on \yrseduc\ did not significantly change: from 1.32 to 1.34. The reason is the follwoing that as it has been already indicated in the problem 5 (a) the sample correlation between \yrseduc\ and \textit{female} is very low, as a result, using formula from the 5 (c) $\tilde{\beta_1} = \hat{\beta_1} + \hat{\beta_2} \delta$ one can conclude that extremely small $\delta$ results in $\tilde{\beta_1} \approx \hat{\beta_1}$ even if $\hat{\beta_2} \neq 0$.
\end{enumerate}
\section*{Problem 8}
Use multiple linear regression of \textit{ahe} on \yrseduc\ and \textit{ba} (a binary variable equaling one for
those who attained a bachelor’s degree) to estimate the marginal value of a bachelor’s degree,
more specifically, the value of going to school for 16 years and receiving a bachelor’s degree,
relative to going to school for 16 years but not receiving a bachelor’s degree. Construct a 95\%
confidence interval for the marginal value of a bachelor’s degree.
\end{document}

>>>>

\begin{center}
	Econometrics 1. Problem Set 2.
\end{center}
\section*{Problem 1}
Suppose that the population model determining $y$ is
\begin{align*}
y = \beta_0 + \beta_1 x_1 + \beta_2x_2 + \beta_3x_3 + u
\end{align*}
and this model satisfies the Gauss-Markov assumptions. However, we estimate the model that
omits $x_3$. Let $\tilde{\beta_1}, \tilde{\beta_2}$ be the OLS estimators from the regression of $y$ on $x_1$ and $x_2$. Show
that the expected value of $\tilde{\beta_1}$ (given the values of the independent variables in the sample) is
\begin{align*}
\expect \tilde{\beta_1} = \beta_1 + \beta_3 \frac{\sum_{i=1}^n \hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align*}
where the $\hat{r_{i1}}$ are the OLS residuals from the regression of $x_1$ on $x_2$.

Following hint, plugging $y_i = \beta_0 + \beta_1x_{i1} + \beta_2x_{i2}+\beta_3x_{i3}$ into
\begin{align*}
\tilde{\beta_1} = \frac{\sum_{i=1}^n \hat{r_{i1}}y_i}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align*}
one can obtain:
\begin{align} \label{eq1}
\tilde{\beta_1} = \frac{\sum_{i=1}^n \hat{r_{i1}}(\beta_0 + \beta_1x_{i1} + \beta_2x_{i2}+\beta_3x_{i3} + u_i)}{\sum_{i=1}^n \hat{r_{i1}}^2} = \nonumber \\
=\frac{\beta_0 \sum_{i=1}^n \hat{r_{i1}} + \beta_1 \sum_{i=1}^n \hat{r_{i1}}x_{i1} + \beta_2 \sum_{i=1}^n \hat{r_{i1}}x_{i2} + \beta_2 \sum_{i=1}^n \hat{r_{i1}}x_{i2} + \sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align}
By properties of OLS estimators (corollary from FOC): $\sum_{i=1}^n \hat{r_{i1}} = 0, \sum_{i=1}^n \hat{r_{i1}}x_{i2} = 0$. Moreover,
\begin{align*}
\sum_{i=1}^n \hat{r_{i1}}x_{i1} = \sum_{i=1}^n \hat{r_{i1}} (\hat{\alpha_0} + \hat{\alpha_1}x_{i2} + \hat{r_{i1}}) = 0 + 0 + \sum_{i=1}^n \hat{r_{i1}}^2
\end{align*}
where $\hat{\alpha_0}, \hat{\alpha_1}$ are OLS estimates for coefficients of regression of $x_1$ on $x_2$. Plugging it into (\ref{eq1}):
\begin{align*}
\tilde{\beta_1} = \beta_1 + \beta_3\frac{\sum_{i=1}^n\hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2} + \frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2}\\
\expect \tilde{\beta_1} = \beta_1 + \beta_3\frac{\sum_{i=1}^n\hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2} + \expect\left[ \frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2} \right]
\end{align*}
\begin{align*}
\hat{r_{i1}} = x_{i1} - \hat{x_{i1}} = x_{i1} - \hat{\alpha_0} - \hat{\alpha_1}x_{i2}\ \to \expect[f(\hat{r_{i1}})|x_{i1}, x_{i2}] = f(\hat{r_{i1}}) \ \forall\ f(\cdot)
\end{align*}
Then,
\begin{align*}
\expect\left[\frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2} \right] = \expect\left[\expect\left[\frac{\sum_{i=1}^n \hat{r_{i1}}u_i}{\sum_{i=1}^n \hat{r_{i1}}^2}\bigg|x_{1}, x_{2}\right] \right] \underset{\text{i.i.d.}}{\propto} \expect\left[\sum_{i=1}^n\hat{r_{i1}}\expect [u_i|x_{i1}, x_{i2}]\right] \underset{\text{G.-M. assump.}}{=} 0 
\end{align*}
Thus, 
\begin{align*}
\expect \tilde{\beta_1} = \beta_1 + \beta_3\frac{\sum_{i=1}^n\hat{r_{i1}}x_{i3}}{\sum_{i=1}^n \hat{r_{i1}}^2}
\end{align*}
\section*{Problem 2}
\begin{enumerate}[(i)]
	\item Consider the simple regression model $y = \beta_0 +\beta_1x+u$ under the first four Gauss-Markov
	assumptions. For some function $g(x)$, for example $g(x) = x^2$ or $g(x) = \ln (1 + x^2)$, define
	$z_i = g(x_i)$. Define a slope estimator as
	\begin{align*}
	\tilde{\beta_1} = \frac{\sum_{i=1}^n (z_i - \bar{z})y_i}{\sum_{i=1}^n (z_i - \bar{z})x_i}
	\end{align*}
	Show that $\tilde{\beta_1}$ is linear and unbiased.
	\item Add the homoskedasticity assumption, MLR.5. Show that
	\begin{align*}
	Var \tilde{\beta_1} = \sigma^2 \frac{\sum_{i=1}^n (z_i -  \bar{z})^2}{\left(\sum_{i=1}^n (z_i - \bar{z})x_i\right)^2}
	\end{align*}
	\item Show directly that, under the Gauss-Markov assumptions, $Var\hat{\beta_1} \le Var\tilde{\beta_1}$, where $\hat{\beta_1}$ is the OLS estimator.
\end{enumerate}

@@

\begin{enumerate}[(a)]
	\item If $\exists\ i: x_i \neq 0$ then the estimator is well-defined.
	\item \begin{align*}
	\expect[\tilde{\gamma}|x_1, \dots, x_n] = \expect\left[\frac{\summa (\gamma x_i + \varepsilon_i)x_i^3}{\summa x_i^4}\bigg|x_1, \dots, x_n\right] &= \gamma + \frac{1}{\summa x_i^4} \summa x_i^3\expect[\varepsilon_i|x_1, \dots, x_n] =\\
	 &\underset{\text{G.-M. assump}}{=} \gamma + 0
	\end{align*}
	\item \begin{align*}
	Var(\tilde{\gamma}|x_1, \dots, x_n) = Var\left[\frac{\summa (\gamma x_i + \varepsilon_i)x_i^3}{\summa x_i^4}\bigg|x_1, \dots, x_n\right] \underset{\text{i.i.d}}{=} \frac{\summa x_i^6 Var[\varepsilon_i|x_i]}{\left(\summa x_i^4\right)^2} 
	\end{align*}
	If $Var(\varepsilon|x) = const = \sigma^2$ then
	\begin{align*}
	Var(\tilde{\gamma}|x_1, \dots, x_2) = \sigma^2 \frac{\summa x_i^6}{\left(\summa x_i^4\right)^2} 
	\end{align*}
\end{enumerate}
\section*{Problem 5}
In this problem set you continue to explore the returns to education and the gender gap
in earnings. The data set is the same as for problem set 1 (cps99 ps1.csv; see problem set 1
for a description of the variables). Consider the regression of average hourly earnings (\textit{ahe}) on
years of education (\textit{yrseduc}). Consider the following three variables that have been omitted
from the regression:
\begin{enumerate}[(a)]
\item gender
\item a binary variable that = 1 if the worker’s last name falls in the first half of the alphabet,
and = 0 if it falls in the second half.
\item the worker’s native ability (for example IQ or some better measure)
\end{enumerate}
For each: Will this omission arguably lead to omitted variable bias? Why or why not?
If your answer is that it will, state the sign of

@@

\textbf{Solution}
\begin{enumerate}[(a)]
	\item Since the education is a sequential process i.e. to get an additional year of university one should primarily finish a school. That means that in order to estimate the effect of $i$-th additional year at school we should randomly choose two groups of children randomly in the sense that each child in the USA should be equally likelly chosen in these groups. Then these children should be randomly distributed among schools in the USA and study there $i-1$ years. Then children from the first group should be expelled from school and finish their education. Children from the second group should be allowed to study one more additional year and after that should also be expelled from the school. Then at some age of those people we should measure average salaries in both groups and compare these means using some statistical tests. 
	\item This protocol actually eliminates omitted variable bias because it eliminates correlation between additional year of education and any other factors which can influence earnings. To be more precise, let assume that earnings are described as follows $y = \beta_0 + \beta_1 \text{\textit{yrseduc}} + \beta_2 \text{\textit{intel.ability}} + u$. But in our controlled randomized experiment intelectual abilities of people and years of education are independent variables. Hence if we somehow measured intelectual abilities and regress them on \textit{yrseduc} in our experiment we would get something like on the Fig \ref{fig2}. 
		\begin{figure}[h]
		\centering
		\includegraphics[width=0.8\textwidth]{iqyrs}
		\caption{}\label{fig2}
	\end{figure}
i.e. estimated coefficient $\delta$ would be close to zero. But estimated coefficient from regression $y$ on \textit{yrseduc} is equal to $\hat{\beta_1} + \delta \hat{\beta_2}$ where $\hat{\beta_1}, \hat{\beta_2}$ are OLS estimates on hypothetical regression $y$ on \textit{yrseduc} and intelectual abilities. THe same thing with any other facctors which can in principle affect earnings. Thus, estimated value of additional school year becomes unbiased. Of course, such kind of experiment is not possible in reality because of ethical issues.
\end{enumerate}
\section*{Problem 7}
Estimate two regressions: (i) \textit{ahe} on \textit{yrseduc} and (ii) \textit{ahe} on \textit{yrseduc} and \textit{female}.
\begin{enumerate}[(a)]
\item In regression (ii), what is the coefficient on \yrseduc? Explain what this means.
\item In regression (ii), test the hypothesis that the population coefficient on female in
specification (ii) is zero, against the hypothesis that it is nonzero, at the 5\% significance level.
In everyday words (not statistical terms), what precisely is the hypothesis is that you are
testing?
\item Does the coefficient on \textit{yrseduc} change from regressions (i) to (ii) in a substantively
important way, that is, is the difference between the two estimates large in a real-world sense?
What is the reason that including \textit{female} in the regression results in a large (or only a small)
change in the coefficient on \yrseduc? (Hint: what is the correlation between \yrseduc\ and
\textit{female}?)

@@

\begin{enumerate}[(a)]
	\item Coefficient on \yrseduc\ is equal to 1.34147. That means that holding a gender of the person fixed each additional year of education on average gives 1.34147 dollars of increase of the per hour earnings.
	\item Using coeftest() it can be obtained that the null hypothesis should be rejected on the 5\% significance level. The hypothesis was that holding a number of years of education fixed both males and females get on average equal salary per hour.
	\item Coefficient on \yrseduc\ did not significantly change: from 1.32 to 1.34. The reason is the follwoing that as it has been already indicated in the problem 5 (a) the sample correlation between \yrseduc\ and \textit{female} is very low, as a result, using formula from the 5 (c) $\tilde{\beta_1} = \hat{\beta_1} + \hat{\beta_2} \delta$ one can conclude that extremely small $\delta$ results in $\tilde{\beta_1} \approx \hat{\beta_1}$ even if $\hat{\beta_2} \neq 0$.
\end{enumerate}
\section*{Problem 8}
Use multiple linear regression of \textit{ahe} on \yrseduc\ and \textit{ba} (a binary variable equaling one for
those who attained a bachelor’s degree) to estimate the marginal value of a bachelor’s degree,
more specifically, the value of going to school for 16 years and receiving a bachelor’s degree,
relative to going to school for 16 years but not receiving a bachelor’s degree. Construct a 95\%
confidence interval for the marginal value of a bachelor’s degree.

The summary of regression is depicted below.
	\begin{figure}[h]
	\centering
	\includegraphics[width=0.8\textwidth]{ba}
	\caption{}\label{fig1}
\end{figure}

Using confint() one can get the following 95\% confidence interval for the marginal value of a bachelor degree:
\begin{align*}
(-0.1332442; 1.705304)
\end{align*}
This interval include 0, hence we cannot infer that for people whose number of years of education are equal, obtainment of bachelor degree on average leads to a higher salary.
\end{document}
